---
title: "Expected Rush Yards"
author: "Lucca Ferraz, Savan Patel, Ryan Coombs, Peyton Stevenson"
format: 
  html:
    toc: true
    toc-location: left
    toc-depth: 3
    number-sections: true
    embed-resources: true
cache: true
---

## Read in Data

```{r}
# loading in data and libraries
library(tidyverse)
library(janitor)
train_data <- read_csv("train_data.csv")
test_data <- read_csv("test_data.csv")
submission_sample <- read_csv("submission_sample.csv")
```

## Cleaning Data

The first step in our process was to standardize coordinates so all players are moving from left to right, as well as identifying the rusher, which team was on offense, and how many yards away from the end zone the offense is.

```{r}
clean_data <- train_data |> 
  clean_names() |> 
  rename(o = orientation) |> 
  mutate(
    # make all plays go from left to right
    x = ifelse(play_direction == "left", 120 - x, x),
    y = ifelse(play_direction == "left", 160 / 3 - y, y),
    # flip player direction and orientation
    dir = ifelse(play_direction == "left", dir + 180, dir),
    dir = ifelse(dir > 360, dir - 360, dir),
    o = ifelse(play_direction == "left", o + 180, o),
    o = ifelse(o > 360, o - 360, o)
  ) |> 
  mutate(is_rusher = ifelse(nfl_id == nfl_id_rusher, 1, 0), # identify rushers
         yards_from_home = ifelse(possession_team == field_position, yard_line , 50 + (50 - yard_line)), # Yards from own goal if PossessionTeam==FieldPosition
         offense_team = ifelse(possession_team == home_team_abbr, 'home', 'away'), # home or away team on offense if PossessionTeam == HomeTeamAbbr
         offense = ifelse(team == offense_team,1,0), # offensive team or defensive team if team == TeamOffense
         yards_to_td = 100 - yards_from_home, # calculate yards to go
  ) |> 
  mutate(x_end = 2*sin(dir*pi/180) + x,
         y_end = 2*cos(dir*pi/180) + y)
```

## Feature Engineering

For the purposes of our model, we engineer various different features. We keep game context variables such as score differential, down and distance, quarter, and offensive/defensive formations/personnel. We also keep features describing the rusher, such as their location, speed, acceleration, and orientation. We devise defender features describing how close the closest defender is, how far the farthest defender is, as well as how far the 25th, 50th, and 75th percentile distance defenders are from the ballcarrier. We also compute how many defenders are "ahead" of the ball, by measuring how many defensive players there are within a 30 degree cone formed from the ball carriers' direction. Once we have that, we measure how close the closest player within the "cone" is, as well as how far the farthest is. We then compute blocking features (how many blockers, how close they are to the ballcarrier, how far apart horizontally and vertically they are to each other). Lastly, we compute convex hull features, such as the total area the offense occupies (calculated using the shoelace formula with the different players as coordinates), as well as the defensive area (calculated the same way). We take these features both for the entire field as well as for the 30 degree cone ahead of the ballcarrier.

```{r, warning = FALSE}
# helper to compute polygon area (useful for features of space control by offense and defense)
poly_area <- function(x, y) {
  if(length(x) < 3) return(NA_real_) # need at least 3 points
  # order points by convex hull
  hull_idx <- chull(x, y)
  x_h <- x[hull_idx]
  y_h <- y[hull_idx]
  # shoelace formula
  0.5 * abs(sum(x_h * dplyr::lead(y_h, default = y_h[1])) -
              sum(y_h * dplyr::lead(x_h, default = x_h[1])))
}

# function to compute all features given the play frame
make_features <- function(play_frame) {
  # play_frame: one tibble of all players at handoff for a single play
  # identify who the rusher is
  rusher_id <- unique(play_frame$nfl_id_rusher)
  rusher <- play_frame |> filter(nfl_id == rusher_id)
  
  # game context features
  context_feats <- tibble(
    game_id = play_frame$game_id[1],
    id = play_frame$id[1],            # take first row safely
    down = play_frame$down[1],
    distance = play_frame$distance[1],
    yard_line = play_frame$yard_line[1],
    quarter = play_frame$quarter[1],
    home_score_before_play = play_frame$home_score_before_play[1],
    visitor_score_before_play = play_frame$visitor_score_before_play[1],
    defenders_in_the_box = play_frame$defenders_in_the_box[1],
    offense_formation = play_frame$offense_formation[1],
    offense_personnel = play_frame$offense_personnel[1],
    defense_personnel = play_frame$defense_personnel[1],
    score_diff = play_frame$home_score_before_play[1] - play_frame$visitor_score_before_play[1]
  )
  
  # rusher features
  rusher_feats <- tibble(
    rusher_x = rusher$x[1],
    rusher_y = rusher$y[1],
    rusher_speed = rusher$s[1],
    rusher_accel = rusher$a[1],
    rusher_orientation = rusher$o[1],
    rusher_dir = rusher$dir[1],
    rusher_yardline_dist = rusher$yard_line[1] - rusher$x[1]
  )
  
  # defender features
  defenders <- play_frame |> filter(team != rusher$team)
  
  defenders <- defenders |> 
    mutate(dx = x - rusher$x,
           dy = y - rusher$y,
           dist = sqrt(dx^2 + dy^2),
           angle = atan2(dy, dx) * 180/pi)
  
  # nearest defender distances
  nearest <- defenders |> 
    summarise(min_defender_dist = min(dist),
              mean_defender_dist = mean(dist),
              dist_25 = quantile(dist, 0.25),
              dist_50 = quantile(dist, 0.50),
              dist_75 = quantile(dist, 0.75))
  
  # defenders ahead of rusher (within ±30° cone)
  defenders_ahead <- defenders |> 
    filter(dx > 0, abs(angle) < 30)
  
  cone_feats <- defenders_ahead |> 
    summarise(cone_min_dist = min(dist, na.rm = TRUE),
              cone_mean_dist = mean(dist, na.rm = TRUE),
              cone_count = n())
  
  # blocker features (use OL only for simplicity)
  blockers <- play_frame |> 
    filter(team == rusher$team,
           position %in% c("C","G","T","OT","OG","LT","RT","LG","RG")) |> 
    mutate(dx = x - rusher$x,
           dy = y - rusher$y,
           dist = sqrt(dx^2 + dy^2))
  
  if(nrow(blockers) > 0) {
    blocker_feats <- blockers |> 
      summarise(
        num_blockers = n(), # how many blockers
        mean_blocker_dist = mean(dist), # avg distance of blocker to rusher
        min_blocker_dist = min(dist), # closest distance of blocker to rusher
        max_blocker_dist = max(dist), # farthest distance of blocker to rusher
        line_width = max(y) - min(y),   # spread of OL horizontally
        line_depth = max(x) - min(x),   # how deep OL is relative to LOS
        avg_blocker_orientation = mean(o, na.rm = TRUE) # avg blocker direction
      )
  } else {
    blocker_feats <- tibble(
      num_blockers = 0,
      mean_blocker_dist = NA,
      min_blocker_dist = NA,
      max_blocker_dist = NA,
      line_width = NA,
      line_depth = NA,
      avg_blocker_orientation = NA
    )
  }
  # convex hull features
  offense <- play_frame |> filter(team == rusher$team)
  defense <- play_frame |> filter(team != rusher$team)
  
  offense_area <- poly_area(offense$x, offense$y)
  defense_area <- poly_area(defense$x, defense$y)
  
  hull_feats <- tibble(
    offense_hull_area = offense_area,
    defense_hull_area = defense_area,
    hull_area_ratio = offense_area / defense_area 
  )
  # convex hull features in rushers' forward cone
  # forward cone defined by dx > 0 and abs(angle) < 30°
  offense_cone <- offense |> 
    mutate(dx = x - rusher$x,
           dy = y - rusher$y,
           angle = atan2(dy, dx) * 180/pi) |> 
    filter(dx > 0, abs(angle) < 30)
  
  defense_cone <- defenders_ahead # already filtered
  
  offense_cone_area <- poly_area(offense_cone$x, offense_cone$y)
  defense_cone_area <- poly_area(defense_cone$x, defense_cone$y)
  
  cone_hull_feats <- tibble(
    offense_cone_area = offense_cone_area,
    defense_cone_area = defense_cone_area,
    cone_area_ratio = offense_cone_area / defense_cone_area
  )
  # combine all features
  bind_cols(context_feats, rusher_feats, blocker_feats, nearest, cone_feats, 
            hull_feats, cone_hull_feats)
}

# calculate features for every play
all_features <- clean_data |> 
  group_by(id) |> 
  group_modify(~ make_features(.x)) |> 
  ungroup()

target_variable <- clean_data |> 
  group_by(id) |> 
  summarise(yards = first(yards))

# go from -99:99 to 0:198 because xgboost models only accept labels from 0 to number of classes minus 1
target_numeric <- as.numeric(target_variable$yards) + 99

nclass <- 199

# fill NA and infinite values with 0
all_features <- all_features |> 
  mutate(offense_cone_area = ifelse(is.na(offense_cone_area), 0, offense_cone_area),
         cone_area_ratio = ifelse(is.na(cone_area_ratio), 0, cone_area_ratio),
         defense_cone_area = ifelse(is.na(defense_cone_area), 0, defense_cone_area),
         offense_formation = ifelse(is.na(offense_formation), "UNKNOWN", offense_formation),
         defenders_in_the_box = ifelse(is.na(defenders_in_the_box), 0, defenders_in_the_box),
         cone_mean_dist = ifelse(is.na(cone_mean_dist), 0, cone_mean_dist),
         cone_min_dist = ifelse(is.infinite(cone_min_dist), 0, cone_min_dist)) |> 
  select(-game_id, -home_score_before_play, -visitor_score_before_play) |> # unhelpful variables
  mutate(down = as.factor(down),
         quarter = as.factor(quarter)) # convert to factor

model_features <- all_features |> 
  select(-id) # do not want the model to overfit to play id's
```

## Modeling

For our model we use an XGBoost. This allows us to capture complex non-linear linteractions between our features while also optimizing for prediction accuracy. To tune the hyperparameters, we apply a grid search and use 5-fold cross-validation to test which hyperparameters work best.
```{r}
#| eval: false
# Identify categorical columns
cat_cols <- sapply(model_features, function(x) is.factor(x) || is.character(x))

# turn categorical columns into factors
model_features[cat_cols] <- lapply(model_features[cat_cols], factor)
# save factor levels to ensure matches with test data
factor_levels <- lapply(model_features[cat_cols], levels)

# one-hot encoding for categorical variables
encoded_features <- model.matrix(~ ., data = model_features[, cat_cols])

# combine with numeric features
numeric_features <- model_features[, !cat_cols]
final_features <- cbind(numeric_features, encoded_features)

# remove intercept column as well as target variable
final_features <- final_features[, !colnames(final_features) %in% c("(Intercept)", "yards")]

# define evaluation metric, the Continuous Ranked Probability Score (CRPS)
crps =  function(preds, dtrain,raw_prob=F) {
  labels <- getinfo(dtrain, "label")
  n_samples <- length(labels)
  preds_matrix <- matrix(preds, nrow = n_samples, byrow = TRUE)
  
  # if preds is not raw probabilities, use softmax to convert them to raw probabilities.
  if (raw_prob == F) {
    # Apply softmax to each row
    softmax <- function(x) exp(x) / sum(exp(x))
    preds_prob <- t(apply(preds_matrix, 1, softmax))
    
  } else {
    
    preds_prob = preds_matrix
    
  }
  
  crps_sum <- 0
  # calculate crps scores
  for (i in 1:n_samples) {
    y <- labels[i] - 99  # Convert back to -99 to 99 range
    p <- cumsum(preds_prob[i, ])  # Cumulative probabilities
    
    # print (preds_prob[i, ])
    crps_sample <- sum((p - (y <= (-99:99)))^2) / 199
    crps_sum <- crps_sum + crps_sample
  }
  
  crps <- crps_sum / n_samples
  return(list(metric = "crps", value = crps))
}

# for cross-validation, take only first 500 rows for quicker runtime 
dtrain <- xgb.DMatrix(data = as.matrix(final_features)[1:500,], label = target_numeric[1:500])


# parameter grid for tuning
grid <- expand.grid(
  nrounds = seq(100, 200, 10),
  eta = c(0.05, 0.1, 0.2),
  max_depth = c(4, 5, 6)
)

results <- list()

# for all possible hyperparameter combos, train model using cross-validation and report test error
# this cross-validation takes over 3 hours to run
for (i in 1:nrow(grid)) {
  params <- list(
    objective = "multi:softprob",
    num_class = nclass,
    eta = grid$eta[i],
    max_depth = grid$max_depth[i]
  )
  
  cv <- xgb.cv(data = dtrain,
               nrounds = grid$nrounds[i],
               eta = grid$eta[i],
               max_depth = grid$max_depth[i],
               # verbose = FALSE,
               num_class = nclass,
               eval_metric = crps,
               early_stopping_rounds = 20,
               maximize = FALSE,
               nfold = 5,
               objective = 'multi:softprob'
               
  )
  
  results[[i]] <- data.frame(
    nrounds = grid$nrounds[i],
    eta = grid$eta[i],
    max_depth = grid$max_depth[i],
    best_iter = cv$best_iteration,
    best_score = min(cv$evaluation_log$test_crps_mean, na.rm = TRUE)
  )
}

# get results
results_df <- do.call(rbind, results)

# best parameters
best <- results_df[which.min(results_df$best_score), ]

# now we can train model using all data
dtrain <- xgb.DMatrix(data = as.matrix(final_features), label = target_numeric)

# train model using best hyperparameters (this also takes over an hour)
xg_model <- xgboost(data = dtrain,
                    nrounds = best$nrounds,
                    eta = best$eta,
                    max_depth = best$max_depth,
                    # verbose = FALSE,
                    num_class = nclass,
                    eval_metric = crps,
                    # nfold = 5,
                    objective = 'multi:softprob'
                    )

# see most important features
library(vip)
vip(xg_model, num_features = 100)
```

## Apply Model to Test Data

```{r}
#| eval: false
# clean test data same way as train
clean_test_data <- test_data |> 
  clean_names() |> 
  rename(o = orientation) |> 
  mutate(
    # make all plays go from left to right
    x = ifelse(play_direction == "left", 120 - x, x),
    y = ifelse(play_direction == "left", 160 / 3 - y, y),
    # flip player direction and orientation
    dir = ifelse(play_direction == "left", dir + 180, dir),
    dir = ifelse(dir > 360, dir - 360, dir),
    o = ifelse(play_direction == "left", o + 180, o),
    o = ifelse(o > 360, o - 360, o)
  ) |> 
  mutate(is_rusher = ifelse(nfl_id == nfl_id_rusher, 1, 0), # identify rushers
         yards_from_home = ifelse(possession_team == field_position, yard_line , 50 + (50 - yard_line)), # Yards from own goal if PossessionTeam==FieldPosition
         offense_team = ifelse(possession_team == home_team_abbr, 'home', 'away'), # home or away team on offense if PossessionTeam == HomeTeamAbbr
         offense = ifelse(team == offense_team,1,0), # offensive team or defensive team if team == TeamOffense
         yards_to_td = 100 - yards_from_home, # calculate yards to go
  ) |> 
  mutate(x_end = 2*sin(dir*pi/180) + x,
         y_end = 2*cos(dir*pi/180) + y)

# compute features for test data
all_test_features <- clean_test_data |> 
  group_by(id) |> 
  group_modify(~ make_features(.x)) |> 
  ungroup()

# fill in NA and infinite values
all_test_features <- all_test_features |> 
  mutate(offense_cone_area = ifelse(is.na(offense_cone_area), 0, offense_cone_area),
         cone_area_ratio = ifelse(is.na(cone_area_ratio), 0, cone_area_ratio),
         defense_cone_area = ifelse(is.na(defense_cone_area), 0, defense_cone_area),
         offense_formation = ifelse(is.na(offense_formation), "UNKNOWN", offense_formation),
         defenders_in_the_box = ifelse(is.na(defenders_in_the_box), 0, defenders_in_the_box),
         cone_mean_dist = ifelse(is.na(cone_mean_dist), 0, cone_mean_dist),
         cone_min_dist = ifelse(is.infinite(cone_min_dist), 0, cone_min_dist)) |> 
  select(-game_id, -home_score_before_play, -visitor_score_before_play) |> 
  mutate(down = as.factor(down),
         quarter = as.factor(quarter)) # change down and quarter to factors

model_test_features <- all_test_features |> 
  select(-id) # remove play id

# identify categorical columns
cat_cols_test <- sapply(model_test_features, function(x) is.factor(x) || is.character(x))

# force same factor levels as train (test data does not contain all formations seen in train)
for (col in names(factor_levels)) {
  # set factor levels
  model_test_features[[col]] <- factor(model_test_features[[col]],
                                       levels = factor_levels[[col]])
  # replace NAs (unseen levels) with first level or baseline
  model_test_features[[col]][is.na(model_test_features[[col]])] <- levels(model_test_features[[col]])[1]
}


# one-hot encoding for categorical variables
encoded_test_features <- model.matrix(~ .,
                                      data = model_test_features[, names(factor_levels)])

# combine with numeric features
numeric_test_features <- model_test_features[, !cat_cols_test]
final_test_features <- cbind(numeric_test_features, encoded_test_features)

# remove intercept column
final_test_features <- final_test_features[, !colnames(final_test_features) %in% c("(Intercept)", "yards")]

# get xgboost matrix for predictions
dtest <- xgb.DMatrix(data = as.matrix(final_test_features))

# make predictions
df_predict <- predict(xg_model, dtest)

# change to matrix
matrix_predict <- matrix(df_predict, nrow = nrow(as.matrix(final_test_features)), byrow = TRUE)

matrix_predict_cumsum <- t(apply(matrix_predict,1,cumsum)) # calculate cumulative probability

# create submission file in same format as example
submission_file <- cbind(data.frame(ID = unique(test_data$ID)),matrix_predict_cumsum)
colnames(submission_file)[2:ncol(submission_file)]= paste0('Yards',-99:99)

# write_csv(submission_file, "submission_file.csv") write to csv
```